Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling

### 2014-12-06 Documents/Paper/ms.tex
More writing intro

where $L(\theta)$ is the likelihood function, and the expectation is taken
with respect to the ``power posterior'' distributions which are intermediate
between the prior $\pi(\theta)$ and the posterior:

\begin{eqnarray}
p_\beta(\theta) \propto \pi(\theta)L(\theta)^\beta.
\end{eqnarray}

If the posterior distribution is a mixture of a slab and a spike,
then for some range of inverse temperatures $\beta$ the probability mass in
the slab and the spike will be comparable. At these temperatures the MCMC
method will be unable to mix and will give an incorrect estimate of
$\left<\log\left[L(\theta)\right]\right>_\beta$.

Nested Sampling \citep{skilling} overcomes these problems by working with an
alternative sequence of probability distributions, proportional to the prior
but with a hard constraint on the value of the likelihood:
\begin{eqnarray}
p_{l}(\theta) \propto \pi(\theta)\mathds{1}\left[L(\theta) > l\right]
\end{eqnarray}
The normalising constants of these distributions are given by
\begin{eqnarray}
X(l) &=& \int \pi(\theta)\mathds{1}\left[L(\theta) > l\right] \, d\theta
\end{eqnarray}
which is the amount of prior mass that has likelihood greater than
$l$. As Nested Sampling proceeds, $X$ shrinks geometrically with time, so
the number of iterations required to compress from the slab to the spike is
of order $\log(V_{\rm slab}/V_{\rm spike})$.

\subsection{Diffusive Nested Sampling}
Diffusive Nested Sampling
\citep[DNS][]{dnest} replaces the posterior distribution with an alternative target
distribution composed of a mixture of the prior distribution with other, more
constrained versions of the prior, facilitating mixing between the multiple modes
and along degeneracy curves.
This is similar to the idea of annealed importance sampling
\citep{neal} or parallel tempering \citep{pt},
where the target distribution is modified from the posterior to something
``easier''. However,
the sequence of distributions used in Nested Sampling avoids some of the
problems with annealing-based methods. In particular, there is no need to choose
a temperature schedule (a potentially large set of tuning parameters),
and the method does not fail when phase transitions
are present.
Figure~\ref{fig:challenges} shows a mock-up of a two dimensional posterior
density with three modes, one of which contains a strong dependence, and
another of which contains a phase transition.

>>>>

where $L(\theta)$ is the likelihood function, and the expectation is taken
with respect to the ``thermal'' distributions which are intermediate
between the prior $\pi(\theta)$ and the posterior:
\begin{eqnarray}
p_\beta(\theta) \propto \pi(\theta)L(\theta)^\beta.\label{eq:annealing}
\end{eqnarray}
If the posterior distribution is a mixture of a slab and a spike,
then for some range of inverse temperatures $\beta$ the probability mass in
the slab and the spike will be comparable. At these temperatures the MCMC
method will be unable to mix and will give an incorrect estimate of
$\left<\log\left[L(\theta)\right]\right>_\beta$.

Nested Sampling \citep{skilling} overcomes these problems by working with an
alternative sequence of probability distributions, proportional to the prior
but with a hard constraint on the value of the likelihood:
\begin{eqnarray}
p_{l}(\theta) \propto \pi(\theta)\mathds{1}\left[L(\theta) > l\right]\label{eq:constrained_prior}
\end{eqnarray}
The normalising constants of these distributions are given by
\begin{eqnarray}
X(l) &=& \int \pi(\theta)\mathds{1}\left[L(\theta) > l\right] \, d\theta
\end{eqnarray}
which is the amount of prior mass that has likelihood greater than
$l$. As Nested Sampling proceeds, $X$ shrinks geometrically with time, so
the number of iterations required to compress from the slab to the spike is
of order $\log(V_{\rm slab}/V_{\rm spike})$. During this process Nested Sampling
visits states with likelihoods intermediate between those of the slab and the
spike, whereas no choice of $\beta$ in Equation~\ref{eq:annealing} will give
these states substantial probability.

\subsection{Diffusive Nested Sampling}
The main difficulty with Nested Sampling is coming up with methods to sample
the constrained prior distributions of Equation~\ref{eq:constrained_prior}.
Diffusive Nested Sampling
\citep[DNS][]{dnest} is an alternative version of Nested Sampling that is
appropriate when MCMC is used to explore the parameter space.

Rather than working with a sequence of constrained prior distributions like in
Equation~\ref{eq:constrained_prior}, DNS
replaces the posterior distribution with an alternative target
distribution composed of a {\it mixture} of the prior distribution with the
constrained priors, facilitating mixing between the multiple modes
and along degeneracy curves.
This is similar to the idea of annealed importance sampling
\citep{neal} or parallel tempering \citep{pt},
where the target distribution is modified from the posterior to something
``easier''. However,
the sequence of distributions used in Nested Sampling avoids some of the
problems with annealing-based methods. In particular, there is no need to choose
a temperature schedule (a potentially large set of tuning parameters),
and the method does not fail when phase transitions
are present.
Figure~\ref{fig:challenges} shows a mock-up of a two dimensional posterior
density with three modes, one of which contains a strong dependence, and
another of which contains a phase transition.
